---
title: "session06 - homework"
format: html
editor: visual
date: 2025-12-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_libraries
#| echo: false

# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesrules)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Introduction

We will be analyzing the data from the following article:

Winter, B., & Matlock, T. (2013). Making judgments based on similarity and proximity. Metaphor and Symbol, 28(4), 219-232.

```{r}
#| label: preprocessing

# Load our datasets
sim <- read_csv("../data/linguistic_similarity.csv")

```

In this study, they investigated the conceptual structure of the metaphor “SIMILARITY IS PROXIMITY.” They did a series of experiments in which they investigated whether participants judge entities to be closer to each other when they are thought to be more similar. They presented participants with descriptions of two cities. These cities either sounded similarly or they sounded differently (`Phon`). Moreover, the descriptions depicted the cities as either semantically similar or different (`Sem`). After reading both descriptions and seeing both names, participant had to draw into a map and had to randomly guess where these two cities are located. The authors measured the `Distance` between the two cities for each participant in millimeters.

# (A) Descriptive statistics and data visualization

Let's visualize the raw `Distance` distribution for both `Sem` categories making a density plot of it and describe the distributions.

```{r}
#| label: show_overall_distribution

# YOUR CODE HERE

sim #just looking at the data first

library(ggplot2)

show_overall_distribution <- 
  ggplot(data=sim, aes(x= Distance, fill = Sem)) +
  geom_density(alpha = 0.5) +
  labs(x= "Distance", y = "Density") +
  theme_minimal()

show_overall_distribution

#get warning for 1 row removed containing non-finite outside the scale; one data point seems to have NA as the distance

#Describing distribution: Both semantically different and semantically similar cities have "peaks" around 75 mm, indicating that regardless of the condition, most people placed the cities at a similar distance from each other. We can also see that the distribution is a bit different for the conditions: for semantically different cities, the answers are more focused around the 75 mm mark, with a few more answers at a greater distance (>120) than semantically similar cities, and fewer answers at a distance shorter than 75mm. For semantically similar cities however, there are more answers in the shorter range, with a gradual rise in answers up until the 75 mm mark, and a sharp drop off after. There is a small spike of answers around the 160 mm mark, which is similar to the different condition.
```

# (B) Think about priors

The researchers are interested in whether semantic similarity between the cities affects how close/far apart participants draw the cities, which we can model with the following formula:

```{r}
#| label: model_formula

mdl_formula = bf(Distance ~ Sem)

```

Use the `get_prior()` function and add `mdl_formula` as its first argument and the `sim` dataset as its second. This will give you a list of parameters that the model will estimate and will need a prior for.

```{r}
#| label: get_prior

get_prior(mdl_formula, sim) 

```

Generally, every coefficient of our model corresponds to one parameter under the class `b`, except for the `Intercept` which is its own class. So we have a prior for the `Intercept` of the linear model and a prior for the coefficient `SemSimilar`. Priors of class `b` for our predictors always receive a flat uniform prior by default if you do not specify it otherwise. The default prior for the `Intercept` is data driven and uses a student-t distributions that covers most of the data. We also have a prior for `sigma` which is the residual variance and again is by default assigned a student-t prior informed by the variance in the data.

Now I want you to specify normally distributed priors which respect the following knowledge we have: Participants needed to draw the cities on a map on a piece of din-A4 paper. The furthest two points on the map where 214 mm apart.

Your task: First use the `plot_normal()` function to find a normal distribution that covers reasonable Distance values under the above described assumptions of the experiment.

Ignore for now that you might generate values that are impossible at the tails of the distribution.

```{r}
#| label: prior_exploration

plot_normal(100, 57)
#logic: 100 is easy as a mid-point and then 57 is halv of 114, so 2sd would lead to max 214? idk if this is wrong but; also ignoring the impossible negative number for now
```

# (C) Specify priors

Now based on your exploration, specify priors. Remember what the two coefficients `Intercept` and `SemSimilar` mean: `Intercept` reflects the average Distance measure for `Sem = Different`, and `SemSimilar` reflects the estimated difference between the Intercept and `Sem = Similar`. Define a prior that assumes normally distributed Distance measures somewhere in the middle of the map for both conditions:

```{r}
#| label: spec_priors

my_priors <- c(
  prior(normal(100, 57), class = Intercept), #what the numbers mean => 100 = where i assume my mean is, normal = normal distribution, 57 = the range of 1 sd, aka accounting for noise around the data
  prior(normal(0, 40), class = b), #note from class: the class b prior you set to 0, you ASSUME null difference between the two semantic conditions, and then plus variation of 50 each way to account for noise
  prior(normal(0, 30), class = sigma)#old notes: i haven't done anything to count the difference between similar and different tho? new notes: and you do not need to, because you are assuming 0 difference, the 40 is just to allow for sampling error/noise/minor variations around 0
) #idk what to do with the sigma values as well?? (new notes: sigma values are assumed to also be 0, with noise of 30) 
#timo in class = did not do the sigma prior because lazy, apparently that is fine to do

```

# (D) Run a prior predictive check

Add your priors to the model below. We will run the model and sample only from our priors, i.e. we ignore the data. This is a good sanity check to see if our priors make sense and what they would predict. Plug in `my_priors` from (C) and run the code chunk.

```{r}
#| label: model_priors

install.packages("cmdstanr") #got an error it can't run on my version of r but seemed to start running anyways? idk

xmdl_prior <- brm(Distance ~ Sem,
            family = gaussian(),
            my_priors,
            sample_prior = "only",
            data = sim,
            seed = 42)


# We do a little wrangling to get it into a nice shape
xmdl_prior |> 
  # first extract the samples
  spread_draws(b_Intercept, b_SemSimilar) |> 
  # Create two columns that represent the two SEM categories
  mutate(Different = b_Intercept,
         Similar = b_Intercept + b_SemSimilar) |> 
  # and then bring into a long format like the data
  select(Different, Similar) |> 
  pivot_longer(everything(), names_to = "Sem", values_to = "Distance") |> 
  # now we plot
  ggplot(aes(x = Distance, fill = Sem)) +
    geom_density() +
    facet_wrap(Sem ~ .)

```

Do the priors reflect what you had in mind? If not, go back to (C) and play around with the priors until you are satisfied.

# (E) Run the full model

If you are happy with your priors, run the full model and plot the posteriors just like in the code chunk above. How did the posteriors change after seeing the data?

```{r}
#| label: model_posteriors

xmdl_full <- brm(Distance ~ Sem,
            family = gaussian(),
            prior = my_priors,
            data = sim,
            seed = 42)

#so r gave up on this a few times with a few error messages, but ran in the end

xmdl_full

#intercept is now  82.55, and semantically similar cities are now -10.13 mm different than sem different cities (aka around 10 mm closer together)
```

After seeing the data the very diffuse priors have given way to much more narrow and shifted posteriors.

We can also plot the data in an alternative way, since we are mostly interested in the difference between the two conditions:

```{r}
#| label: alternative_plot

# We do a little wrangling to get it into a nice shape
xmdl_full |> 
  # first extract the samples for the differences only
  spread_draws(b_SemSimilar) |> 
  # plot
  ggplot(aes(x = b_SemSimilar)) +
    geom_density(fill = "purple") 

```

Now, one last step, before we move on. Let's check whether this model could have generated the data using `pp_check()`.

Throughout our modeling process, we'll constantly use `pp_check()`, which is a way for us to know whether our model could've generated the data. This is a way of investigating whether the model 'fits' well. PP checks can tell us, for example, whether the chosen response distribution (in this case Gaussian) was appropriate, or whether the model will have to be expanded, that is, whether particular predictors are missing.

Describe whether the data (dark line) and the posterior samples (light blue line) match up and if now, what the nature of the mismatch is.

```{r}
#| label: pp_check

pp_check(xmdl_full, ndraws = 100)

# YOUR ASSESSMENT HERE
#honestly that looks almost spot on to what our first graph looked like. the y part looks almost identical to the semantically different graph. the posterior samples (light blue line) are a bit more normally distributed and also a bit less tall. the start of the curve and the end of the curve line up with the y curve, which shows us that they are both keeping within the constraits of the model (between 0 and 214 mm distance). the peak of both curves is also in more or less the same place, although the y peak is higher up on the y axis compared to the mean of the light blue lines. this peak is also around the 75 mm mark, which matches the original data. the biggest mismatch is the curve falling after the peak, where the dark line falls a lot quicker than the light blue line, but this is not a major difference and is likely because we are forcing the light blue line to give us a normal distribution. 

```

# OPTIONAL (F) fix the assumption violation

Can you fix the mismatch between the observed data (dark blue line) and the simulated data (light blue lines) with, for example, a transformation of the Distance measure? Run the model again with the transformed values and do the pp check again.

```{r}
#| label: transform

# YOUR CODE HERE
#square root transformation = gentle transformation for gentle right-side tail; log-transform too aggressive

```

# OPTIONAL (G) specify priors that will reduce the effect of SEM to zero

Now I want you to specify priors that will dominate the posterior, i.e. they will eliminate any predicted difference between conditions.

```{r}
#| label: too-strict_priors

# YOUR CODE HERE

```
