---
title: "session07 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: false
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, lme4, brms, bayestestR)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

### Introduction to the problem

We will be working with Quantese again, see `01_exercise.qmd` and the other data resources for a refresher.

```{r}
#| label: load-quantese

quantese <- read_csv("../data/quantese_naming_sim.csv") 
# add log-transformed reaction time
quantese <- quantese |> 
  mutate(log_rt = log(rt))

# have a look
quantese

```

In an earlier homework, we have assessed the following hypothesis:

*"We hypothesize that log-transformed reaction times are affected by lexical frequency"*

We tried to run the following model that included an appropriate random effect structure but it did not converge with `lme4`:

```{r}
#| label: lmer_quantese

# formulate the corresponding LMER
quant_lmer <- lmer(log_rt ~ log_frequency +
                     (1 + log_frequency | PartID) +
                     (1 | ItemID), 
                    data = quantese)
# your receive a warning: "boundary (singular) fit: see help('isSingular')"

```

Now we will re-run this model within the Bayesian framework:

### (A) Specify priors

Check which priors this model needs and specify weakly informative priors.

Remember that weakly informative priors are priors that do not bias the posteriors in any direction, but reasonably constrain the parameter space, i.e. only reasonable values are assumed.

For priors of class `Intercept` and `b`, constrain the parameter space with a wide normal distribution.

```{r}
#| label: get_prior

# YOUR CODE HERE

mdl_formula = bf(log_rt ~ log_frequency)

get_prior(mdl_formula, quantese) #ask timo abt this again because i genuinely do not understand how and what i am supposed to be reading here

#reasonable values here would be above 0?
summary(quantese)
#according to the data, which i do not tink i am supposed to be using rn but bear with me, log_freq and log_rt have a max of 6 and 7.8 respectively, so the range of 0 to 10 should be fine? i am not a math person

plot_normal(5, 5)
??plot_normal

library(bayesrules)
plot_normal(10, 5) #this feels wrong but i do not have the knowledge to articulate why, or what would be better, but since the most important part is nothing below 0 i think we're good? maybe?

# specify weakly informative priors
my_priors <- c(
  prior(normal(10, 5), class = Intercept),
  prior(normal(0, 30), class = b))

```

We can also now get used to specifying priors for classes `sd` and `cor`. Priors of class `sd` are priors for the variance associated with our random effects. You see here 2 `sd` priors for group `ItemID`. The first one is just a generic prior for all `ItemID` coefficients and the second one with the `coef = Intercept` stands for the random intercept variation between items. Similarly for group `PartID`, we have 3 priors. The first one is again just a generic prior for all `PartID` coefficients. The second one with the `coef = Intercept` stands for the random intercept variation between participants. And the last one with the `coef = log_frequency` stands for the random slope variation between participants for the predictor `log_frequency`. How do we specify these priors. We have usually no idea what to expect from these parameters, so we here also specify very diffuse priors centered on zero. We pick a cauchy distribution (remember? super squashed normal distributions) because they have thick tails. Let's make them super diffuse and pick a `sd = 3` for both items and participants.

```{r}
#| label: add_sd_priors

# add weakly informative priors for sd
my_priors <- c(my_priors, 
               prior(cauchy(0, 3), class = 'sd', group = 'ItemID'),
               prior(cauchy(0, 3), class = 'sd', group = 'PartID'))

```

And finally, if your model as random slopes, they have by default a correlation parameter for intercept and slope. We are not getting into the math here and will follow advice from smart people. We use Lewandowski-Kurowicka-Joe priors (shape = 1): `lkj(1)`. Specifying this prior for random slope models actually helps a lot with the fitting procedure.

```{r}
#| label: add_cor_priors

# add weakly informative priors for cor
my_priors <- c(my_priors, 
               prior(lkj(1), class = 'cor'))

```

### (B) Run model

Now run the model with 4 chains, 1000 warm-up and 1000 post-warm-up samples per chain. Check that chains are reasonably mixed and that sampling yielded a sufficient effective sample size. If not address the issues appropriately (check slides again).

```{r}
#| label: model_quant

# if the model takes way too long to sample (more than 5-10 mins), use the following subset
# quantese_sub <- quantese |> 
#   filter(PartID %in% c("01","02","03","04",
#                        "05","06","07","08"))

# YOUR CODE HERE
library(brms)
install.packages("cmdstanr")
install.packages("cmdstanr", repos = c("https://mc-stan.org/r-packages/", getOption("repos")))
library(cmdstanr)
install_cmdstan(cores = 2)
cmdstanr::check_cmdstan_toolchain(fix = TRUE)


model_quant <- brm(log_rt ~ log_frequency +
                     (1 + log_frequency | PartID) +
                     (1 | ItemID),
            family = gaussian(),
            prior = my_priors,
            data = quantese,
            chains = 4,
            cores = 4, 
            iter = 2000,
            warmup = 1000,
            thin = 1,
            seed = 42)

model_quant

?mcmc_plot
mcmc_plot(model_quant, type = 'trace') #notif: No divergences to plot. 7 plots show up but very small and i can't figure out how to make them bigger, but they all look sufficiently caterpillar-y, so i will say yes to convergence

?mcmc_dens_overlay
mcmc_dens_overlay(model_quant, pars = 'b_Intercept') #idek but it cannot find the function despite me finding it in search so idk
library(bayesplot) #ok now it worked
#close enough imo
model_quant

# Rhat values are ~ 1.00 = 1.01, which unsure if that is just within potential lack of convergence, or just outside of it? a bit unclear from the powerpoint
# bulk ESS > 400 = check for log_frequency, but not the case for Intercept (bulk ESS 289)
# tail ESS > 100 = check
#unsure if i should maybe run again

model_quant2 <- brm(log_rt ~ log_frequency +
                     (1 + log_frequency | PartID) +
                     (1 | ItemID),
            family = gaussian(),
            prior = my_priors,
            data = quantese,
            chains = 4,
            cores = 4, 
            iter = 6000,
            warmup = 4000,
            thin = 1,
            seed = 42)


mcmc_plot(model_quant2, type = 'trace') #similar to above

mcmc_dens_overlay(model_quant2, pars = 'b_Intercept') #looks better than earlier, more overlap


model_quant2

# Rhat values are ~ 1.00 = 1.01 again, still unsure if that is just within potential lack of convergence, or just outside of it
# bulk ESS > 400 = check for both
# tail ESS > 100 = check for both
#definitely better than the last one
#google says rhat 1.01 is at the upper boundary of "good" and i might come back to try and fix it, but will accept that and keep going for now
#oh wait i am blind? the first rhat value i 1.00 (ItemID, sd(Intercept))
#in PartID we have rhat 1.01 for sd Intercept, 1.00 for sd log_frequency, and 1.00 for correlation, and then 1.o1 for regression coefficients and 1.00 for sigma values
#happy with that
```

### (C) Plot posterior predictions

Find a way to extract posterior samples from the model for the `Intercept` and the `log_frequency` coefficient and plot the predicted posterior mean linear model alongside its 95% Credible Intervals. There are many different ways to get there.

```{r}
#| label: plot_predictions

# YOUR CODE HERE

fitted.values(model_quant2) #that does not look usable

post_draw <- as_draws_df(model_quant2)
post_draw
colnames(post_draw)

library(bayesplot)
mcmc_areas(post_draw, pars = c("b_Intercept", "b_log_frequency"), prob = 0.95) #what am i looking at right now, this does not look at all correct
#i unfortunately do not have the time to start figuring this out so this is how far i got, sorry timo TT
```
