---
title: "session08 - homework"
format: html
editor: visual
date: 2025-07-17
author: Timo Roettger
execute:
  error: false
  warning: false
  message: false
  cache: true
bibliography: ../resources/bibliography.bib
---

# Preamble: Loading packages and configuration

```{r}
#| label: data_and_libraries
#| echo: false

# just run this code chunk
# function to ignoring the setting of the relative path below when knitting
run_if_not_knitting <- function(expr) {
  if (!isTRUE(getOption("knitr.in.progress"))) {
    eval(expr)
  }
}

# nifty code using the pacman package
# it checks if the packages specified below are installed, if not, they will be installed, if yes, they will be loaded
if (!require("pacman")) install.packages("pacman")
pacman::p_load(rstudioapi, tidyverse, tidybayes, brms, bayesplot, bayestestR, modelr)

# set the current working directory to the one where this file is
run_if_not_knitting(current_working_dir <- dirname(rstudioapi::getActiveDocumentContext()$path))
run_if_not_knitting(setwd(current_working_dir))

# options to increase efficiency of brms models (optional)
#options(brms.backend = "cmdstanr")
options(mc.cores = parallel::detectCores())

```

# Exercise

We will be re-analyzing the data from @winter2013making.

```{r}
#| label: preprocessing_sim

# Load our datasets
sim <- read_csv("../data/linguistic_similarity.csv") |> 
  mutate(Sem = as.factor(Sem),
         Phon = as.factor(Phon))
sim

```

We will now retrace the original article's complete analysis. In their original analysis, they were interested in whether (a) semantic similarity between cities, (b) phonological similarity between cities, and (c) their interaction affects how close/far apart participants draw the cities.

## (a) Specify Priors

Specify the model formula and appropriate weakly informative priors for all relevant predictors. Contrast-code the predictors.

```{r}
#| label: model_formula_sim

my_priors <- #YOUR PRIORS HERE

  model_formula <- bf(Distance~Sem*Phon) #note that i figured out this was incorrect later down, not gonna pretend i didn't mess up at first

get_prior(model_formula, sim) 

my_priors <- c(
  prior(normal(100, 50), class = Intercept), #map was around 214? so estimating ish half the map in between
  prior(normal(0, 50), class = b)) #expecting 0 difference between semsim and semdif
#leaving sigma out since timo said the default is usually fine, might come back and check if something goes wrong

contrasts(sim$Sem)<- contr.sum(2)
contrasts(sim$Phon)<- contr.sum(2)

```

## (b) Run the appropriate model for the research question

```{r}
#| label: model_sim

#YOUR MODEL HERE

model1 <- brm(model_formula, data = sim, prior = my_priors)
#ok so that did not run lol

library(cmdstanr)

brm(model_formula, data = sim, prior = my_priors, backend = "cmdstanr") #ok so since i only see sem1 i likely messed up trying to get phon in here

model_formula2 <- bf(Distance~Sem + Phon + Sem*Phon)

model_yay <- brm(model_formula2, data = sim, prior = my_priors, backend = "cmdstanr") #there we go, that's more like it
#i feel rlly stupid for not doing this from the beginning but we move
```

## (c) Posterior predicive check

Check the overall model fit with posterior predictive checks. If the model does not fit the data well, fix the problem (check `06_homework_solutions.qmd` for inspiration).

```{r}
#| label: model_fit

#YOUR CODE HERE
pp_check(model_yay, ndraws = 100)
#the peak is very close to where the peak is, just much higher, and the bump isn't there because we specified normal distribution, it looks alright
#update reran and it looked much better so honestly will keep if the rest looks good

model_yay #rhat 1.00 for everything slay, bulk ess and tail ess also looking nice and high

library("bayesplot")

mcmc_plot(model_yay, type = 'trace') #oh those look very nice and healthy yay


```

## (d) Assess sampling diagnostics

Check that chains are reasonably mixed and that sampling yielded a sufficient number of effective sample size. If there are issues, address them accordingly.

```{r}
#| label: sampling_diagnostics

# YOUR CODE HERE

summary(model_yay)
#i mean: rhat is 1.00; bulk_ess and tail_ess are both high, as we saw in the plots above the chains agree and mix well, so idk what more to write in this section? have i missed a part on how to do this?
```

## (e) Interpret the results

Now interpret the model relative to the research question and describe uncertainty, visualize posteriors, calculate posterior probabilities of direction, and calculate Bayes Factors for evidence for and against the posterior coefficients being point zero.

(Note: When you want to extract interaction terms like `b_Sem1:Phon1`, you have to put single quotation marks around them because R struggles with interpreting the colon)

```{r}
#| label: interpret_sim

#YOUR CODE HERE
print(model_yay)
#wait i may have messed up, i am not sure what my intercept is 
get_prior(model_yay)
#i cannot tell if sem1 and phon1 are the different or similar conditions :c
#based on the model output and alphabetical order, it is tempting to conclude that the intercept is the mean of the semantically different conditions, and sem1 and phon1 are the semantically similar conditions
#which would make people place semantically similar cities further apart than semantically different cities, while placing phonetically similar cities closer to each other, with a small interaction between the semantic and phonetic conditions.
#i feel like i've messed up already but idk how
#we have very high sigma values, so a lot of variation in responses

```

## (g) Write up your entire analysis:

YOUR WRITE UP HERE

## OPTIONAL (h) compare your interpretation to the original findings

YOUR COMPARISON HERE

# References

<!-- References will auto-populate in the refs div below -->

::: {#refs}
:::
